---
title: "EFA ANALYSIS"
author: "Lucy Gitiria"
output: html_document
---
##  PACKAGES NEEDED
```{r, eval=False,warning = FALSE}
library(readxl)
library(lavaan)
library(psych)
library(tidyverse)
library(polycor)
library(kableExtra)
library(apaTables)
library(corrplot)
library(semTools) # for additional functions in SEM
library(semPlot) # for path diagram
```

#  DATA
```{r}
NSD2687_1_no_1 <- read_excel("D:/THESIS 2022/2018 Data/Studiebarometeret 2018, faggruppe/NSD2687-1-no 1.xlsx")
```
#  DATA PREPARATION
```{r, eval=FALSE}
Data<-NSD2687_1_no_1
preli.df<-Data[c(6,13,14,93:102)]
apply(preli.df[1:13],2,table,exclude=NULL)#A glance at the subset data
preli.df[1:13][preli.df[1:13]==9999]<-NA # change 9999 to NAs
preli.df[1:13][preli.df[1:13]==999]<-NA # change 999 to NAs

```
#  SUBSET 6 STUDY PROGRAMS
```{r, eval=FALSE}
AKADM<-preli.df[preli.df$Utd_type == 'Ã˜KADM',]   # business and adminstration-Enterprising
DATAIT<-preli.df[preli.df$Utd_type == 'DATA-IT',] # Conventional
KUNST<-preli.df[preli.df$Utd_type == 'KUNST',]    # Artistic
SIVING<-preli.df[preli.df$Utd_type == 'SIVING',]  # Investigative
SYKEPLEIE<-preli.df[preli.df$Utd_type == 'SYKEPLEIE',] #Social
TEKN.FAG<-preli.df[preli.df$Utd_type == 'TEKN-FAG',]   #Realistic

# New Dataset
#Put all data frames into list
list.dfprel <- list(SIVING, TEKN.FAG, KUNST,SYKEPLEIE,AKADM,DATAIT)

#Merge all data frames in list
ANALYSIS.DATApreli<-Reduce(function(x, y) merge(x, y, all=TRUE), list.dfprel)
#Remove NAs
DF.ANALYSIS<-ANALYSIS.DATApreli[complete.cases(ANALYSIS.DATApreli[1:13]),]
```

##  DESCRIPTIVE STATISTICS
```{r}
describe(DF.ANALYSIS)
```
```{r}
#Subset LO items
data1<-DF.ANALYSIS[c(4:13)]
apply(data1[1:10],2,table,exclude=NULL)#A glance at the subset data
```
# **EXPLORATORY FACTOR ANALYSIS**
## EFA Data
```{r}
EFA_DATA<-data1
```
## Data Check
EFA is a multivariate procedure that necessitates that the multidimensional position of each data point be considered.
A group of measured variables might exhibit univariate normality and yet be multivariate non-normal. That is, the joint distribution of all the variables might be non-normal. The psych package offers an implementation of Mardia's multivariate tests (1970) and an accompanying Q–Q plot for multivariate checks. MVN package can check both uni and multivariate normality once.

*Multivariate,Univariate Normality* 
```{r, include=TRUE}
library(MVN)
# Multivariate,Univariate Normality and Descriptives
result <- mvn(data= EFA_DATA, mvnTest="mardia",  univariateTest="AD")
result
#plot using kable
kable(result$univariateNormality, booktabs=TRUE, format="markdown") #Not normal
#Box plots from graphics package
library(graphics)
boxplot(EFA_DATA,notch=FALSE, boxfill= "royalblue", whiskcol="blue",pch=16,outcol="firebrick")
#Histograms to illustrate 
kable(mvn(data= EFA_DATA,univariatePlot="histogram"), booktabs=TRUE, format="markdown")# command for all in one
```
Skew > 2.0 or kurtosis > 7.0 would indicate severe univariate nonnormality (Curran et al., 1996). These univariate and multivariate statistics seem to indicate that all measured variables are non-normally distributeds so there should be much concerns about correlations being restricted due to variable distributions. 
NOTE.Skew (departures from symmetry) and kurtosis (distributions with heavier or lighter tails and higher or flatter peaks) of all variables seem to be close to normal.

*Bivariate characteristics of the data*
```{r,include=TRUE}

# Bivariate characteristics of the data #psych package (Revelle, 2019)
pairs.panels(EFA_DATA, stars = TRUE) 

```
Correlations moderate and pretty close, data not normally distributed,linear relation exhibited.
Use polychoric correlation matrix beacause the data is not normally distributed.
```{r,warning = FALSE, eval=FALSE}
attach (EFA_DATA)
names(EFA_DATA)
#Correlation matrix
cor(EFA_DATA) #pearsons
```
*Compare correlation output with polychoric results below.*

## **EFA using polychoric -using psych package**
```{r,warning = FALSE }
#Calculated Polychoric Correlation
EFA_correl <- mixedCor(EFA_DATA, c=NULL, p=1:10)
str(EFA_correl)
```
```{r}
## Getting the correlation matrix of the dataset.
EFA_polychoric <- EFA_correl$rho
# Explore the correlation structure of the dataset.
corrplot(EFA_polychoric, method = "number", order = "hclust")
```

*Suitability of data for Factor Analysis*
To see if it’s ok to do EFA with the data, calculate Kaiser Meyer Olkin test (KMO) Bartlett’s test of sphericity. 
```{r}
#Using psych package (Revelle, 2019)
#KMO test
KMO(EFA_polychoric) 
```
KMO results indicate that the Overall MSA = 0.89 is close to 1 indicating it is meritorious.
```{r}
#Bartlett's test for sphericity
cortest.bartlett(EFA_polychoric,n=6758)
```
Bartelett's test has a p value of 0- whic is below .05, thus reject the null-proceed with factor analysis.

*Factor Extraction with Principal Axis Factoring ()PAF*
```{r}
# Use PAF on EFA_polychoric
EFA_correl_pa <- fa(EFA_polychoric, nfactors =, fm = "pa")
# Sort the communalities of EFA_correl_pa.
sort(EFA_correl_pa$communality, decreasing = TRUE) 

```
```{r}
# Sort the uniqueness of hsq_correl_pa.
sort(EFA_correl_pa$uniqueness, decreasing = TRUE)
```

*Determining the number of factors*
Use the correlation matrix, EFA_polychoric, computed based on the initial dataset.
```{r}
## Check out the scree test and the Kaiser-Guttman criterion.
scree(EFA_polychoric)
```

```{r}
# Use parallel analysis for estimation with the minres extraction method.
fa.parallel(EFA_polychoric, n.obs = 6758, fm = "minres", fa = "fa", main = "Parallel Analysis Scree Plot")
```
- Keep one factor
*Factor rotation*
```{r}
# EFA with 1 factors.
f_EFA <- fa(EFA_polychoric, nfactors = 1)
```

```{r}
# Check the default rotation method.
f_EFA$rotation
```
```{r}
# Oblimin rotation with 1 factors.
f_EFA_oblimin <- fa(EFA_polychoric, nfactors = 1, rotate = "oblimin")
f_EFA_oblimin
```

```{r}
# Check the factor loadings.
print(f_EFA_oblimin$loadings, cut=0.4)
```
```{r}
# Create the path diagram of the latent factors.
fa.diagram(f_EFA_oblimin)
```
# Remove fagsspes and selvst- have communalities below .4
```{r}
uten_fgs_ogSelv<-EFA_DATA[c(1,2,3,5:9)]
```

```{r}
#Calculated Polychoric Correlation
EFA_correl8items <- mixedCor(uten_fgs_ogSelv, c=NULL, p=1:8)
#str(EFA_correl8items)
```
```{r}
## Getting the correlation matrix of the dataset.
EFA_polychoric8items <- EFA_correl8items$rho
# Explore the correlation structure of the dataset.
corrplot(EFA_polychoric8items, method = "number", order = "hclust")
```

```{r}
# Oblimin rotation with 1 factors.
f_EFA_oblimin8items <- fa(EFA_polychoric8items, nfactors = 1, rotate = "oblimin")
f_EFA_oblimin8items
```
*Proportion Var 0.48, high residuals-RMSR= 1.0->not the best fit* 

```{r}
# Check the factor loadings.
print(f_EFA_oblimin8items$loadings, cut=0.4)
```

```{r}
# Create the path diagram of the latent factors.
fa.diagram(f_EFA_oblimin8items)
```



**EFA WITH TWO FACTORS**
```{r}
# Oblimin rotation with 2 factors.
f_EFA_oblimin2 <- fa(EFA_polychoric, nfactors = 2, rotate = "oblimin")
f_EFA_oblimin2
```
*Cumulative Var 56%, rmsr=.04*

```{r}
# Check the factor loadings.
print(f_EFA_oblimin2$loadings, cut=0.4)
```

```{r}
# Create the path diagram of the latent factors.
fa.diagram(f_EFA_oblimin2)
```

**EFA WITH THREE FACTORS**
```{r}
# Oblimin rotation with 3 factors.
f_EFA_oblimin3 <- fa(EFA_polychoric, nfactors = 3, rotate = "oblimin")
f_EFA_oblimin3
```

```{r}
# Check the factor loadings.
print(f_EFA_oblimin3$loadings, cut=0.35)
```

```{r}
# Create the path diagram of the latent factors.
fa.diagram(f_EFA_oblimin3)
```






